GEE models

### 2017-10-20 doc/submission/reviewer-response-2.md

    We thank the reviewer for these supportive comments and hope that our responses below
    adequately address the issues and suggestions raised.

- *"A key question for me is the choice of using the GEE approach, which is of
course an established and appropriate method for many research questions, but I
think could be better justified here. The main advantages I am aware of are:
Richer correlation structures, easily handles missing data, robust standard
errors. Here, the number of follow up visits is not large and the amount of
missing data is relatively small. My first thought when approaching the analysis
would therefore be to consider a standard linear regression analysis with
robust/clustered standard errors to incorporate repeated measures, and potential
multiple imputation of missing values and a comparison with complete case
analysis. So it would be nice to briefly spell out the things that GEE offers
here over other approaches to make this clear (in the ESM if need be so as not
to add to the word count). Currently the main statement is the ease of handling
missing data - which is handy but perhaps not essential - and some comments on
the working correlation matrix, which are not terribly informative. So some
statements about the improvement in estimates under a specific correlation
structure that is not easily handled otherwise, for instance, might be helpful.
Of note is that in the ESM it is stated that "other ... matrices had similar fit" -
it may be worth noting that the model fit may be similar, but estimates may be
more efficient under a particular correlation structure."*

    We agree entirely with the reviewer's suggestion to more fully describe the justification for using
    GEE over other methods. In the revised manuscript, we have expanded our rationale
    for using GEE in both the Methods and the ESM Methods sections. We chose the GEE
    technique for a variety of reasons. As the reviewer states, GEE is able to
    handle visit dates that differ between participants (e.g. some participants
    come exactly 3 years later for a follow up while others come some months
    before or after 3 years) and for dealing with the inherent correlational
    structure underlying longitudinal data. Because of these features, among
    others, GEE can compute robust standard errors of the estimates that more
    accurately reflect the underlying data. The ability to specify a working
    correlation matrix makes this a powerful technique for longitudinal data
    compared to standard linear regression, as it removes the underlying
    assumption that observations are independent (which is not the case with
    longitudinal data). While choosing the specific correlation matrix is
    important, GEE still produces robust estimates and standard errors even with
    a misspecified correlation matrix. In the case of longitudinal data, an
    auto-regressive correlation matrix is the most appropriate, statistically and
    biologically, since measurements on an individual over time will change (e.g.
    people gain or lose weight over time) but still be highly correlated to their 
    previous measurement (e.g. people don't gain or lose too much weight from their
    original weight).

@@

    The reviewer brings up excellent points. The reason NEFA were included as a
    proportion in the PLS analysis rather than as a concentration was due to the
    inherent interdependence between NEFA values as a proportion. They all must
    equal to 100. It is this dependency structure in the data that allows PLS to
    more accurately identify clusters within the data. Concentration data do
    not have this inherent dependency (all NEFA values do *not* equal to 100).
    If one NEFA increases in concentration, other NEFA values do not necessarily
    decrease, as would be the case with the proportion. We have clarified this in
    the methods.

    The reason we did not include all NEFA variables (as proportion or as
    concentration) into a single GEE model is because GEE cannot handle this
    level of complexity particular due to the collinearity between NEFA leading
    to biased estimates and model instability. How the reviewer
    described setting the variables in the regression equation may be useful for
    some research questions and with fewer variables. However, we had 22
    different fatty acids (44 if including mol% and nmol/ml) and 4 different
    outcomes. The high dimensionality of this analysis makes interpretation
    using standard linear regression approaches extremely challenging, not to
    mention introducing potential bias due to correlation among the NEFA
    (especially as a proportion). These reasons are why we chose to run multiple
    GEE models for each NEFA-outcome pairing in addition to running a PLS
    analysis with all NEFA in a single model (since PLS is designed for this
    particular analysis).

    The reviewer is correct that proportions *alone* are not enough to draw a
    meaningful conclusion, as per how the reviewer described it. This, we
    believe, is a major strength of our study since previous literature using other
    fatty acid fractions (e.g. phospholipid) generally only investigate
    proportions but do not report on or measure the concentration. Concentration
    data are more technically challenging to quantify in the lab and so it is understandable
    why large cohort studies do not measure these data. Using both concentration
    and proportion, we can identify what specific role the absolute size and
    specific proportion of individual NEFA have on the risk for diabetes.

@@

    - *"List of abbreviations: if this is to be an exhaustive list many are not
    given, OGTT, HOMA, NEFA"*

        These have been added as suggested.

    - *"QIC - presumably this is quite like the AIC and other information criteria -
    maybe mention this, and how the calculation differs? Is the QIC specifically for
    GEE models? Also in the ESM: "This identifies a specific model from a number of
    models that is the 'best' fit and least complex of all the models" - it can’t be
    the best fit AND least complex, as mentioned in the previous sentence it’s a
    trade-off - so maybe omit this sentence."*

        The reviewer is correct that QIC is similar to AIC. We have expanded on
        the explanation about QIC vs other information criterion. The reviewer
        is correct that QIC is specific to GEE-type models. Because GEE uses
        quasi-likelihood rather than maximum likelihood estimation (as other
        techniques use such as mixed effects models), information criteria such
        as AIC cannot be used since they rely on maximum likelihood. We agree
        that the wording surrounding 'fit' and 'complex' could be clearer.
        However, we disagree that it should be omitted since techniques such as
        QIC *do* select models that have both a good fit and are less complex.
        While techniques such as AIC are powerful model selection methods, they
        have an inherent limitation in that models with more parameters will
        have a better fit. So models with more parameters (more 'complex') will be
        more likely to be selected by default when using AIC. QIC and other
        similar information criteria techniques on the other hand try to balance
        this limitation by more strongly penalizing a more
        'complex' model (more parameters) while still emphasizing a good fit. We
        have clarified this in the manuscript with the wording: "... that has
        both a good model fit while keeping model complexity low (less variables
        included)." See the reference below for more information on QIC:

@@

    3. *"Please add the number of observations for each clinic visit in Table 1."*

        As requested, this has been added lto Table 1.

    4. *"Please add the number of incident and prevalent diabetes cases to the
    respective exclusions in ESM Figure 1. For example: n=736 had baseline data
    collected, n=XX diabetes cases were excluded; n=619 had 3-year data, n=XX
    incident diabetes cases were excluded..."*

        The exact number of excluded cases of diabetes is dependent on this analysis
        and doesn't reflect the main PROMISE cohort. Since the CONSORT diagram 
        represents the main cohort, we included the number of excluded diabetes
        cases in the text of the Results section as the text reflects the 
        present analysis. (See first paragraph of Results section.)

@@

    - *"With respect to peripheral insulin sensitivity (referred to in the
    discussion) the effects of saturated fatty acids may strongly depend on
    metabolic phenotype, ie more pronounced in insulin resistant subjects, was
    initial metabolic phenotype taken into account in the present analysis?"*

        The initial metabolic phenotype and subsequent influence of saturated fatty
        acids is taken into account in the time-by-fatty acid interaction. In
        the present analysis, there were no significant time-by-NEFA
        interactions.

    - *"it is a pity that the analysis of fatty acids composition in
    phospholipids and serum cholesteryl esters have been published in a separate
    paper, altogether this would add very much to the total interpretation."*

        While we agree with the reviewer that inclusion of the phospholipid and
        cholesteryl ester (as well as the triacylglycerol) fractions would have
        added to the interpretation, the volume and complexity of the analyses
        would not have been publishable given current journal word and figure
        count limits. This was the reason we decided to publish the fractions in
        separate manuscripts, to maintain a clear and focused narrative within
        the constraints of the journals limits. If interested, the reference is:

        Johnston LW, Harris SB, Retnakaran R, et al (2016) Longitudinal
        associations of phospholipid and cholesteryl ester fatty acids with
        disorders underlying diabetes. J Clin Endocrinol Metab 101:2536–2544.

### 2017-10-20 doc/esm.Rmd
removed gen png for esm
minor typo fix in esm caption

```{r fig-captions, include=FALSE}
knitr::opts_chunk$set(dev = "png")

@@

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit for the main PROMISE cohort in addition of the sample size determination for the present analysis.')
sfig_dag_isi <- sfig("dag_isi", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISI. Node and arrow description: Green with a triangle is the exposure; blue with a line/I is the outcome; empty blue is an ancestor of the outcome; empty green is an ancestor of the exposure; and empty red is an ancestor of both the outcome and the exposure; green arrow is on causal pathway; red arrow is a biasing pathway; black arrow is not along the pathway from the exposure to the outcome. DAGitty recommended a minimal adjustment of waist, sex, and MET (physical activity).")
sfig_dag_issi2 <- sfig("dag_issi2", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISSI-2. Node and arrow description: Green with a triangle is the exposure; blue with a line/I is the outcome; empty blue is an ancestor of the outcome; empty green is an ancestor of the exposure; and empty red is an ancestor of both the outcome and the exposure; green arrow is on causal pathway; red arrow is a biasing pathway; black arrow is not along the pathway from the exposure to the outcome. DAGitty recommended a minimal adjustment of waist, ethnicity, and MET (physical activity).")
sfig_corr_conc <- sfig('corr_conc', 'Pearson correlation heatmap of non-esterified fatty acids (*nmol/mL*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_corr_pct <- sfig('corr_pct', 'Pearson correlation heatmap of non-esterified fatty acids (*mol%*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted, *excluding waist circumference*. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
```

>>>>

```{r fig-captions, include=FALSE}

@@

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit for the main PROMISE cohort in addition of the sample size determination for the present analysis.')
sfig_dag_isi <- sfig("dag_isi", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISI. Node and arrow description: Green with a triangle is the exposure; blue with a line/I is the outcome; empty blue is an ancestor of the outcome; empty green is an ancestor of the exposure; and empty red is an ancestor of both the outcome and the exposure; green arrow is on causal pathway; red arrow is a biasing pathway; black arrow is not along the pathway from the exposure to the outcome. DAGitty recommended a minimal adjustment of waist, sex, and MET (physical activity).")
sfig_dag_issi2 <- sfig("dag_issi2", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISSI-2. Node and arrow description: Green with a triangle is the exposure; blue with a line/I is the outcome; empty blue is an ancestor of the outcome; empty green is an ancestor of the exposure; and empty red is an ancestor of both the outcome and the exposure; green arrow is on causal pathway; red arrow is a biasing pathway; black arrow is not along the pathway from the exposure to the outcome. DAGitty recommended a minimal adjustment of waist, ethnicity, and MET (physical activity).")
sfig_corr_conc <- sfig('corr_conc', 'Pearson correlation heatmap of non-esterified fatty acids (*nmol/mL*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_corr_pct <- sfig('corr_pct', 'Pearson correlation heatmap of non-esterified fatty acids (*mol%*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted, *excluding waist circumference*. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
```
